A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-08-22 main.tex
4.1 - variable selection table results description

We begin by comparing the ordinary lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) in various settings and broken down into bias and variance components.  Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods due to the fact that 10 additional predictors have been added to the model, but none of them actually explain any variability in the outcome. The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the naive lasso. This improvement is more substantial in the independent subpopulations simulated data than the admixed simulated data. \anna{quantify this for all 3 data settings using $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2$} This is due to the fact that the discrete, nonoverlapping genetic subpopulations more directly correspond to the distinct fixed environmental effects than do the more continuous and blended subpopulations in the admixture data. In other words, a greater amount of the environmental effect is in the column space of the projection matrix of $\bX$ for the independent subpopulations data than in the admixed data. As such, there is a greater amount of confounding in the independent subpopulations which can be corrected by the PC-Lasso and LMM-Lasso methods.

@@

\anna{var selection results}
\begin{table}[H]
\centering
\begin{tabularx}{400pt}{cclYYY}
\toprule
$\eta$ & $\xi$ & \multicolumn{1}{c}{Method} & 1D Linear Admixture & Independent Subpopulations & Semi-Empirical \\ 
\midrule
\multirow{6}{*}{0.5} & \multirow{3}{*}{0} & Lasso & 0.25 (0.05) & 0.23 (0.05) & 0.37 (0.06) \\ 
& & PC-Lasso & 0.23 (0.05) & 0.21 (0.05) & 0.36 (0.06) \\ 
& & LMM-Lasso & 0.25 (0.05) & 0.23 (0.05) & 0.41 (0.06) \\
\cmidrule{2-6}
& \multirow{3}{*}{0.8} & Lasso & 0.23 (0.05) & 0.25 (0.05) & 0.29 (0.06) \\ 
& & PC-Lasso & 0.28 (0.05) & 0.36 (0.06) & 0.47 (0.06) \\ 
& & LMM-Lasso & 0.29 (0.05) & 0.39 (0.06) & 0.58 (0.06) \\ 
\bottomrule
\end{tabularx}
\label{tab:var_sel}
\caption{True positive rates, mean (SD), for different data types in the presence and absence of environmental confounding with coarse subpopulation structure and dichotomous-discordant environmental effects.}
\end{table}

>>>>

We begin by comparing the ordinary lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) and broken down into bias and variance components. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods due to the fact that 10 additional predictors have been added to the model, but none of them actually explain any variability in the outcome. The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the naive lasso. This improvement is more substantial in the independent subpopulations simulated data than the admixed simulated data. \anna{quantify this for all 3 data settings using $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2$} This is due to the fact that the discrete, nonoverlapping genetic subpopulations more directly correspond to the distinct fixed environmental effects than do the more continuous and blended subpopulations in the admixture data. In other words, a greater amount of the environmental effect is in the column space of the projection matrix of $\bX$ for the independent subpopulations data than in the admixed data. As such, there is a greater amount of confounding and potential estimation bias in the independent subpopulations which can be corrected by the PC-Lasso and LMM-Lasso methods.

@@

In addition to MSE, these methods' performance in terms of variable selection accuracy were quantified using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Table \ref{tab:var_sel} displays the mean true positive rate and corresponding standard deviation in the same settings as those illustrated in Figure \ref{fig:mse}. The TPRs echo the MSE results in that Lasso and LMM-Lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-Lasso and PC-Lasso achieve substantial improvements in TPR over Lasso across data types. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Semi-Empirical data sets than in the 1D Linear Admixture data. Again, this is due to the reduced proportion of environmental confounding that is in the column space of $\boldsymbol{P}_{\bX}$, which results in a lesser degree of confounding of the SNP-phenotype associations.

\begin{table}[H]
\centering
\begin{tabularx}{400pt}{cclYYY}
\toprule
$\eta$ & $\xi$ & \multicolumn{1}{c}{Method} & 1D Linear Admixture & Independent Subpopulations & Semi-Empirical \\ 
\midrule
\multirow{6}{*}{0.5} & \multirow{3}{*}{0} & Lasso & 0.25 (0.05) & 0.23 (0.05) & 0.37 (0.06) \\ 
& & PC-Lasso & 0.23 (0.05) & 0.21 (0.05) & 0.36 (0.06) \\ 
& & LMM-Lasso & 0.25 (0.05) & 0.23 (0.05) & 0.41 (0.06) \\
\cmidrule{2-6}
& \multirow{3}{*}{0.8} & Lasso & 0.23 (0.05) & 0.25 (0.05) & 0.29 (0.06) \\ 
& & PC-Lasso & 0.28 (0.05) & 0.36 (0.06) & 0.47 (0.06) \\ 
& & LMM-Lasso & 0.29 (0.05) & 0.39 (0.06) & 0.58 (0.06) \\ 
\bottomrule
\end{tabularx}
\caption{True positive rates, mean (SD), for different data types in the presence and absence of environmental confounding with coarse subpopulation structure and dichotomous-discordant environmental effects.}
\label{tab:var_sel}
\end{table}
